---
title: "Cytometry Packages In Bioconductor"
format: html
---

# Background

The purpose of this .qmd file is to help extract the list of current Bioconductor packages that contain the "FlowCytometry" biocView, allowing for the subsequent evaluation of their use metrics. 

# Getting Stated

## Determining Current Flow Cytometry Packages

Attach the required R packages to your local environment via the `library()` call. 

```{r}
# BiocManager::install("BiocPkgTools")
library(BiocPkgTools)
library(dplyr)
```

Load data about the current packages
```{r}
Packages <- biocPkgList()
# colnames(Packages)
```

```{r}
FlowCytometryPackages <- Packages |>
     filter(sapply(biocViews, function(v) "FlowCytometry" %in% v)) |>
     rename("PublishDate" = 'Date/Publication') #|> arrange(desc(PublishDate))
```

Write out the main column to outputs for use by the other scripts. 

```{r}
# colnames(FlowCytometryPackages)
ToCSV <- FlowCytometryPackages |> select(Package)
StoreHere <- file.path("outputs", "BioconductorPackages.csv")
write.csv(ToCSV, StoreHere, row.names=FALSE)
```

## Current Packages Years Bioconductor

Load the required R packages

```{r}
library(dplyr)
library(stringr)
library(xml2)
library(rvest)
```

```{r}
BioconductorPackages <- file.path("outputs", "BioconductorPackages.csv")
FlowCytometryPackages  <- read.csv(BioconductorPackages, check.names=FALSE)
packages <- FlowCytometryPackages |> pull(Package)
```

Load in the data scapper function

```{r}
#| eval: TRUE

DataScrapper <- function(x){

file <- x

Page <- paste0("https://www.bioconductor.org/packages/release/bioc/html/", file, ".html")

page <- read_html(Page)
text <- page  %>% html_element("body") %>% 
  html_element("main div.container") %>% html_text2()

Imports <- str_extract(text, "Imports\t[^\t]+")  # grab everything after "Imports\t" until next tab
imports_text <- str_remove(Imports, "^Imports\t")
#imports_vector <- str_split(imports_text, ",")[[1]] %>% str_trim()
#imports_vector 

license_line <- str_extract(text, "License\t[^\t]+")
license <- str_remove(license_line, "^License\t") %>% str_trim()
since_line <- str_extract(text, "In Bioconductor since\t[^\t]+")
since <- str_remove(since_line, "^In Bioconductor since\t") %>% str_trim()

Data <- data.frame(package=file, since=since, license=license, imports=imports_text)
return(Data)
}
```

Parse the dataset and save out the .csv

```{r}
#| eval: TRUE
AllFlow <- purrr::map(.x=packages, .f=DataScrapper) |> bind_rows()

cfg <- yaml::read_yaml("https://bioconductor.org/config.yaml")
CurrentVersion <-  cfg$release_version

AllFlow <- AllFlow |> mutate(version = CurrentVersion)

Timing <- AllFlow |> 
  mutate(
    since = str_extract(since, "\\([^()]+\\)$") |>  
            str_remove_all("[()]") |>              
            str_remove(" years") |>               
            as.numeric(),                          
    since = if_else(is.na(since), 0, since)      
  )

Today <- Sys.Date()
filename <- paste0("CurrentFlowPackages_", Today, ".csv")

StoreHere <- file.path("outputs", "CurrentFlowPackages", filename)

write.csv(Timing, StoreHere, row.names=FALSE)
```

## Deprecated Packages

Load required packages

```{r}
library(rvest)
library(xml2)
library(dplyr)
library(stringr)
library(purrr)
```

Define Removed Packages Page

```{r}
DeprecatedURL <- read_html("https://bioconductor.org/about/removed-packages/")
uls <- html_elements(DeprecatedURL, "ul.inline_list")
```

Load Helper Scrapper

```{r}
DataScrapperDeprecated <- function(uls, target_version){
uls_version <- uls[
  map_lgl(uls, ~ {
    hrefs <- html_attr(html_elements(.x, "a"), "href")
    any(str_detect(hrefs, paste0("^/packages/", target_version, "/")))
  })
]

packages_321 <- map_dfr(
  uls_version,
  ~ {
    links <- html_elements(.x, "a")

    tibble(
      package = html_text(links),
      url     = html_attr(links, "href"),
      version = target_version
    )
  }
)

return(packages_321)
}
```

Update the following as needed manually (Sorry!)
```{r}
CurrentVersion

Versions <- c("3.23", "3.22", "3.21", "3.20", "3.19", "3.18", "3.17", "3.16", "3.15", "3.14", "3.13", "3.11",
"3.10", "3.9", "3.8", "3.7", "3.6", "3.5", "3.4", "3.1", "3.0", "2.14", "2.13", "2.12", "2.11", "2.10",
"2.9", "2.8", "2.7", "2.6", "2.5", "2.4", "2.3", "2.2", "2.1", "2.0", "1.9")
```

```{r}
DeprecatedBioconductor <- purrr::map(.x=Versions, .f=DataScrapperDeprecated, uls=uls) |> bind_rows()

DeprecatedBioconductor <- DeprecatedBioconductor |> mutate(url = paste0("https://bioconductor.org", url))

filename <- "DeprecatedPackages.csv"
StoreHere <- file.path("outputs", filename)

write.csv(DeprecatedBioconductor, StoreHere, row.names=FALSE)
```


```{r}
DataScrapper2 <- function(data, x) {

  internal <- data %>% filter(package %in% x) %>% slice(1)
  Page <- internal %>% pull(url)
  version <- internal %>% pull(version)

  # Normalize URL before reading -------------------------------
  normalize_url <- function(urls) {
    vapply(urls, function(u) {
      if (grepl("\\.html$", u, ignore.case = TRUE)) {
        u
      } else if (grepl("\\.", u)) {
        paste0(sub("(.*)\\.[^\\.]*$", "\\1", u), ".html")
      } else {
        paste0(u, ".html")
      }
    }, character(1))
  }

  Page <- normalize_url(Page)

  # Safe read_html: returns NULL on error ----------------------
  page <- tryCatch(read_html(Page), error = function(e) NULL)

  # If the page failed, bail out now ---------------------------
  if (is.null(page)) {
    return(NULL)
  }

  text <- page %>%
    html_element("body") %>%
    html_element("main div.container") %>%
    html_text2()

  Imports <- str_extract(text, "Imports\t[^\t]+")
  imports_text <- str_remove(Imports, "^Imports\t")

  biocviews_line <- str_extract(text, "(?<=biocViews\\t).*?(?=\\t)")

  license_line <- str_extract(text, "License\t[^\t]+")
  license <- str_remove(license_line, "^License\t") %>% str_trim()

  since_line <- str_extract(text, "In Bioconductor since\t[^\t]+")
  since <- str_remove(since_line, "^In Bioconductor since\t") %>% str_trim()

  Data <- data.frame(
    package   = x,
    since     = since,
    license   = license,
    biocviews = biocviews_line,
    imports   = imports_text,
    version = version
  )

  return(Data)
}

```

```{r}
PackageList <- DeprecatedBioconductor |> pull(package)
PackageList <- PackageList |> unique()
PackageList <- PackageList[!grepl("\\.", PackageList)]

Dataset <- map(.x=PackageList, data=DeprecatedBioconductor, .f=DataScrapper2)
Dataset <- compact(Dataset)
Dataset <- bind_rows(Dataset)

FlowDeprecated <- Dataset |>
  filter(str_detect(biocviews, regex("FlowCytometry", ignore_case = TRUE)))

FlowDeprecated <- FlowDeprecated |>
  filter(!str_detect(biocviews, regex("FlowCytometryData", ignore_case = TRUE)))

FlowDeprecated <- FlowDeprecated |> 
  mutate(
    since = str_extract(since, "\\([^()]+\\)$") |>  
            str_remove_all("[()]") |>              
            str_remove(" years") |>               
            as.numeric(),                       
    since = if_else(is.na(since), 0, since)   
  )

cfg <- yaml::read_yaml("https://bioconductor.org/config.yaml")
CurrentVersion <-  cfg$release_version


FlowDeprecated <- FlowDeprecated %>% mutate(
    decimal = as.numeric(sub(".*\\.", "", version)),
    current_decimal = as.numeric(sub(".*\\.", "", CurrentVersion)),
    years_ago = (current_decimal - decimal) * 0.5) |>
  select(-decimal, -current_decimal)
```

```{r}
Today <- Sys.Date()
filename <- paste0("DeprecatedFlowPackages_", Today, ".csv")
StoreHere <- file.path("outputs", "DeprecatedFlowPackages", filename)

write.csv(FlowDeprecated, StoreHere, row.names=FALSE)
```

## Plotting

### Licenses Pie Chart

```{r}
library(ggplot2)

Timing <- Timing |>
  mutate(
    license = recode(
      license,
      `GPL (>=2)` = "GPL (>= 2)",
      `GPL (>=3)` = "GPL (>= 3)"
    )
  )

Timing <- Timing |>
  mutate(
    license = recode(
      license,
      `GPL-2` = "GPL (>= 2)",
      `GPL-3` = "GPL (>= 3)",
      `GPL-3 + file LICENSE` = "GPL (>= 3)",
    )
  )

Timing |> pull(license) |> unique()



# Count number of packages per license
license_counts <- Timing %>%
  count(license)

# Pie chart
plot <- ggplot(license_counts, aes(x = "", y = n, fill = license)) +
  geom_col(width = 1, color = "white") +   # create bar
  coord_polar(theta = "y") +               # convert to pie
  theme_void() +                            # remove axes
  labs(fill = "License", title = "Bioconductor Packages by License") +
  geom_text(aes(label = n),                 # add counts on slices
            position = position_stack(vjust = 0.5))


plot
```

```{r}

Timing <- Timing |> arrange(desc(since))
ggplot(Timing, aes(x = since, fill = license)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Years Since Original Publication",
    y = "Share of Licenses",
    fill = "License"
  ) +
  theme_minimal()
```

### Timeline
```{r}
library(dplyr)
library(ggplot2)

# Preserve ordering by since for y-axis
AllFlow_plot <- Timing %>%
  arrange(desc(since)) %>%
  mutate(package = factor(package, levels = package))  # keeps longest at top

# Plot horizontal timeline with reversed axis labels
plot <- ggplot(AllFlow_plot, aes(x = 0, xend = since, y = package, yend = package)) +
  geom_segment(size = 2, color = "steelblue") +
  scale_x_continuous(
    trans = "reverse",           # reverse only axis numbers
    breaks = seq(0, max(AllFlow_plot$since), by = 5)  # set breaks as needed
  ) +
  labs(
    x = "Years in Bioconductor",
    y = "Package",
    title = "Duration of Packages in Bioconductor"
  ) +
  theme_minimal()

# Save the plot
if(!dir.exists("images")) dir.create("images")

ggsave(
  filename = "images/Timeline.png",
  plot = plot,
  width = 6,
  height = 10,
  dpi = 600
)

```

```{r}
order <- AllFlow_plot |> pull(package)

Data <- read.csv("data/Dataset.csv", check.names=FALSE)

Data <- Data |> filter(Month %in% "Nov")

Data <- left_join(AllFlow_plot, Data, by="package")

Data$package <- factor(Data$package, levels=order)
```

```{r}

plot <- ggplot(Data, aes(x = since, y = Nb_of_distinct_IPs)) +
  geom_point(aes(color = package), size = 3, alpha = 0.7) +  # each package a colored dot
  labs(
    x = "Years in Bioconductor",
    y = "Number of Downloads",
    title = "Package Downloads vs. Time in Bioconductor",
    color = "Package"
  ) +
  theme_minimal()

plotly::ggplotly(plot)
```


# Dependencies

```{r}
order

AllFlow <- read.csv("data/AllFlowPackages.csv", check.names=FALSE)
AllFlow <- AllFlow |> select(package, imports)

# Split by comma
split_items <- lapply(split_items, function(x) {
  x <- trimws(x)               # remove leading/trailing spaces
  x <- gsub("\\(.*?\\)", "", x) # remove anything in parentheses
  x <- trimws(x)               # trim again in case spaces left
  x
})

filtered_items <- lapply(split_items, function(x) intersect(x, order))
AllFlow$filtered_column <- sapply(filtered_items, paste, collapse = ",")

Hmm <- AllFlow |> select(package, filtered_column)
```

```{r}
Used <- Hmm |> filter(package %in% c("flowCore", "flowWorkspace", "openCyto", "ggcyto", 
"CytoML", "PeacoQC", "flowSOM", "flowGate", "flowAI", "flowCut", "cytoMEM", "tidyFlowCore")) |> pull(package)

Used2 <- Hmm |> filter(package %in% c("flowCore", "flowWorkspace", "openCyto", "ggcyto", 
"CytoML", "PeacoQC", "flowSOM", "flowGate", "flowAI", "flowCut", "cytoMEM", "tidyFlowCore"))

split_items <- strsplit(Used2$filtered_column, ",")

# Trim whitespace (in case) and unlist to make a single vector
all_items <- trimws(unlist(split_items))
Dependencies <- unique(all_items)

CummulativeUsed <- c(Used, Dependencies) |> unique()
CummulativeUsed_sorted <- CummulativeUsed[order(match(CummulativeUsed, order))]
```
